Novelty Detection Using Graphical Models for Semantic Room Classification

### 2011-07-01 semantic-nd/article.tex
English fixes, many left, but I got no time or patience.

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
\author{André Susano Pinto\inst{1}\thanks{Corresponding author: \email{andresusanopinto@gmail.com}} \and Andrzej Pronobis\inst{2} \and Luis Paulo Reis\inst{1}}
%
%\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

@@

The novelty detection algorithm is implemented on a cognitive robot Dora the Explorer,
which already uses a developed architecture based on probabilistic graphical models 
oriented towards dealing with and reasoning about uncertain semantic information~\cite{ijcai}.
One of the major problems with using the probabilistic models of the environments for novelty
detection is selecting the optimal threshold above which the test sample is considered
novel. This problem becomes more difficult when, as in case of Dora, the representation
grows as the robot explores the environment. Methods are required to find
optimal parameters given the current structure of the model which constantly changes. To this end, this work 
studies methods for novelty threshold selection using probabilistic graphical models.

@@

Based on this, it can be said that an optimal novelty detection system is aims at 
defining an order relation on all the possible inputs equivalent
to the order defined by the function: $P(\overline{novel}|x)$.
Then, the detector can be described by the largest $P(\overline{novel}|x)$ accepted which is the principle of thresholding.

@@

% Explain approach of using conditional probability
\subsection{Conditional Probability}
\label{sec:conditional-prob}
The conditional probability models the distribution of variables given that the sample
is not novel for the agent.
This is: it should model the distribution of the sensed features considering that the
knowledge the agent possesses holds true. Under that its natural to model it with a graphical
model that combines the information acquired with the relations that the describe the
relations learned by the agent between the variables.

@@

The factors connecting the variables are trained by the system by searching databases of
common knowledge to build potentials describing how likely it is that a specific set of
values for a set of variables is likely to occur.
For instance: it is very likely to find a cereal box in a kitchen; and it is unlikely to find
bathroom connected to another bathroom.
We propose using such a graph $G$ built by the conceptual layer for modelling $P_G(x)$
as an approximation for $P(x|\overline{novel})$.

@@

Its important to notice that in several cases assuming it to be constant leads to
discarding the factor.
Nonetheless, here the distributions are dynamically changing as the system learns
more on the environment.
So the normalizing argument $P(x)$ has to be evaluated for each new subset of $x$.

@@

Having a graphical model $G$ built to model the known data distribution and a model
$U$ for the unconditional probability a novelty threshold would be given by:
$P_G(x)/P_U(x)$.
Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of
variables $x$ into the same measure units (error rate), such that a static threshold can be implemented.
For example the conditional probability would yield very small values on large sets of variables
$x$ than in small sets due to the spreading over the dimensions of the sample space.
As so a novelty measure is seen as a ratio on how much introducing the known concepts helps to
understand the observed result.

\subsection{Semi Supervised: Using Unlabelled Data}
Nonetheless, its often the case that there is access to extra data that allows to
obtain a better approximation to the unconditional probability than the uniform one.
In specific, all the knowledge of the agent can be considered to hold true apart from
complete knowledge on the categories of a room.
In that case a single big factor can be used to model all the variables directly
dependent on the room which the category cannot be assumed to be known as illustrated in
\autoref{fig:unconditional-prob-graph}.

@@

For practical reasons it is impossible to train such a factor, and simplifications need
to be performed. And here, it was assumed that it can be approximated by factorizing
it in several single factors such that all variables become independent.
Additionally those single factors can easily be trained by using unlabelled data,
obtaining this way a graphical model $I$,
as illustrated in \autoref{fig:single-unconditional-prob-graph},
to be used as approximation for the unconditional probability.

@@

In order to verify the performance of the proposed threshold functions a synthetic dataset
was generated. As the point on this initial work was only to test the correctness of
the presented threshold function and approximation capabilities by using a uniform or
an independent model, only information regarding direct features of a room were
modelled and no structured knowledge such as room connectivity was taken
in account.

The synthetic distribution assumes that an independent and variable number of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 9 different measured feature
types. The number of sensed features is dynamic and mimic the type of information
extracted when running on the robot. Due to that it is possible that on a given sample
a certain feature type can be present more than once or not be present at all
(e.g.: room shape is extracted from 2D laser scans in more than one position in the room,
information about detected objects is only present when the robot previously tried to
detect objects on a given room).

The sensed properties and room categories were chosen to mimic as close as possible
the reality and they are based on a previously build ontology from web data.
There is in total 11 different room categories ranging from: corridor, hallway,
1 person office, 2 person office, bathroom, conference hall, etc\dots, and there is
9 different extracted features: room size, room shape, room appearance and 6 different objects
(e.g. book, cereal box, computer).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}
\label{sec:conclusion}
In this paper we presented how to define a stable novelty threshold function on
top of \emph{probabilistic graphical models} instantiated dynamically from sensed
semantic data.
The presented technique is based on the ratio between a conditional and
unconditional probability and when perfect information exists it performs an optimal
novelty detection threshold under the assumption that $P(novel)$ is constant across
all the graph structures.

>>>>

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
\author{André Susano Pinto\inst{1}\thanks{Email: \email{andresusanopinto@gmail.com}} \and Andrzej Pronobis\inst{2} \and Luis Paulo Reis\inst{1}}
%
%\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

@@

The novelty detection algorithm is implemented on a cognitive robot Dora the Explorer,
which already uses a developed architecture based on probabilistic graphical models 
oriented towards dealing with and reasoning about uncertain semantic information~\cite{ijcai}.
One of the major problems with using the probabilistic models of the environments for novelty
detection is selecting the optimal threshold above which the test sample is considered
novel. This problem becomes more difficult when, as in case of Dora, the representation
grows as the robot explores the environment. Methods are required to find
the right threshold given the current structure of the model which constantly changes.
To this end, this work 
studies methods for novelty threshold selection using probabilistic graphical models.

@@

Based on this, it can be said that a novelty detection system aims at 
defining an order relation on all the possible inputs equivalent
to the order defined by the function: $P(\overline{novel}|x)$.
Then, the detector can be described by the largest $P(\overline{novel}|x)$ accepted which is the principle of thresholding.

@@

% Explain approach of using conditional probability
\subsection{Conditional Probability}
\label{sec:conditional-prob}
The conditional probability models the distribution of variables given that the agent
knowledge holds true and can be used to describe the generating classes of the sensed
sample.
Under that its natural to model it with a graphical model that combines the learned
variables and categories as well with the relations between them.

@@

The factors connecting the variables are trained by the system by searching databases of
common knowledge to build potentials describing how likely it is that a specific set of
values of certain variable types is likely to occur.
For instance: it is very likely to find a cereal box in a kitchen; and it is unlikely to find
bathroom connected to another bathroom.
We propose using such a graph $G$ built by the conceptual layer for modelling $P_G(x)$
as an approximation for $P(x|\overline{novel})$.

@@

Its important to notice that in several cases assuming it to be constant leads to
discarding the factor.
Nonetheless, here the distributions are dynamically growing as the system learns
more on the environment.
So the normalizing argument $P(x)$ has to be evaluated for each new subset of $x$.

@@

Having a graphical model $G$ built to model the known data distribution and a model
$U$ for the unconditional probability a novelty threshold would be given by:
$P_G(x)/P_U(x)$.
Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of
variables $x$ into the same measure units (error rate), such that a static threshold can be implemented.
For example the conditional probability would yield very small values on large sets of variables
$x$ than in small sets due to the spreading over the dimensions of the sample space.
A novelty measure is seen as a ratio on how much introducing the known concepts helps to
understand the observed result.

\subsection{Semi Supervised: Using Unlabelled Data}
Nonetheless, its often the case that there is access to extra data that allows to
obtain a better approximation to the unconditional probability than the uniform one.
In specific, all the knowledge of the agent can be considered to hold true apart from
complete knowledge on the categories of a room.
In that case a single big factor can be used to model all the variables directly
dependent on the possibly novel room as illustrated in
\autoref{fig:unconditional-prob-graph}.

@@

For practical reasons, it is impossible to train such a factor, and simplifications need
to be performed. Here, it was assumed that it can be approximated by factorizing
it in several single factors such that all variables become independent.
Additionally those single factors can easily be trained by using unlabelled data.
Obtaining this way a graphical model $I$,
illustrated in \autoref{fig:single-unconditional-prob-graph},
to be used as approximation for the unconditional probability.

@@

In order to verify the performance of the proposed threshold functions, a synthetic dataset
was generated. As the point on this initial work was only to test the correctness of
the presented threshold function and approximation capabilities by using a uniform or
an independent model, only information regarding direct features of a room were
modelled and no structured knowledge such as room connectivity was taken
in account.

The synthetic distribution assumes that an independent and variable number of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 9 different measured feature
types. The number of sensed features is dynamic and mimic the type of information
extracted when running on the robot. Due to that it is possible that on a given sample
a certain feature type can be present more than once or not be present at all
(e.g.: room shape is extracted from 2D laser scans in more than one position in the room,
and information about detected objects is only present when the robot previously tried to
detect objects on a given room).

The sensed properties and room categories were chosen to mimic as close as possible
the reality and they are based on a previously built ontology from web data.
There is in total 11 different room categories ranging from: corridor, hallway,
1 person office, 2 person office, bathroom, conference hall, etc\dots, and there is
9 different extracted features: room size, room shape, room appearance and 6 different objects
(e.g. book, cereal box, computer).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}
\label{sec:conclusion}
In this paper we presented how to define a stable novelty threshold function on
top of \emph{probabilistic graphical models} instantiated dynamically from sensed
semantic data.
The presented technique is based on the ratio between a conditional and
unconditional probability and when perfect information exists it performs an optimal
novelty detection under the assumption that $P(novel)$ is constant across
all the graph structures.
